Staged Abstract Interpreters

### 2018-11-03 paper/10_intro.tex
bg

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation had been emerged in 1990s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. However, it is still worth 
to investigate the idea deeper from a modern perspective based on meta-programming 
for general abstract interpreters. \note{TR: sounds weak, what is new?}
Practically, precise whole-program analyses formulated as abstract interpretation 
can be very slow due to the inherent complexity of the algorithm and the large scale 
of the analyzed programs \cite{toman_et_al:LIPIcs:2017:7121} \todo{other cite?};
particularly, we would like a systematic approach to optimize program 
analyzers and meanwhile minimally modifying the analyzer programs.
In the present work, we show how to achieve this goal by bring existing 
high-performance computing facilities, in the style of multi-stage generative
programming, to abstract interpretation semantic artifacts, i.e., 
abstract interpreters.

>>>>

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation had been emerged in late 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. However, it is still worth 
to investigate the idea deeper from a modern perspective based on meta-programming 
for general abstract interpreters. \note{TR: sounds weak, what is new?}
Practically, precise whole-program analyses formulated as abstract interpretation 
can be very slow due to the inherent complexity of the algorithm and the large scale 
of the analyzed programs \cite{toman_et_al:LIPIcs:2017:7121} \todo{other cite?};
particularly, we would like a systematic approach to optimize program 
analyzers and meanwhile minimally modifying the analyzer programs.
In the present work, we show how to achieve this goal by bring existing 
high-performance computing facilities, in the style of multi-stage generative
programming, to abstract interpretation semantic artifacts, i.e., 
abstract interpreters.

### 2018-11-03 paper/20_bg.tex
bg

We give the concrete semantics of \TLang by showing a big-step definitional interpreter for it.
But before doing that, let's first see a general interface for interpreters, whether concrete or abstract, 
staged or unstaged. The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Ident},
\texttt{Addr} and etc., which will be instantiated concretely later. The abstract types \texttt{Env} and \texttt{Store}
are required to conform their specifications, \texttt{EnvSpec} and \texttt{StoreSpec}, which define 
the operations such as lookup and update on them, nevertheless the detail of them are omitted.  
The store component is necessary when modeling side effects such as mutations. 
The trait also requires the implementators provide the initial values of environment and store.

@@

Readers may notice an additional abstract type \texttt{R[+T]} is defined in the trait,
which means \texttt{R} represents a higher-kinded type and will used to specify the binding-time, 
i.e., staged or not. Since the LMS framework we are using allows us to specify binding-time purely
based on types, the unstaged higher-kinded type can be simply instantiated as an identity type constructor: 
\texttt{type NoRep[+T] = T}. \todo{where talk about stage poly?}

Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, the address space is infinite thus type \texttt{Int} is used here.
The final returned type \texttt{Ans} is a pair of \texttt{Value} and \texttt{Store}, where
the values are either closure or numbers.

@@

Then we define evaluation function that dispatches differently for different expression.
Note that an allocation function is used when we need to update the store: 
in the concrete case, function \texttt{alloc} always returns a fresh address of the store.

\begin{lstlisting}
object NoRepConcreteEval extends Concrete {
  type R[+T] = NoRep[T]
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case If0(cnd, thn, els) => ...
      case AOp(op, e1, e2) => ...
      case Rec(x, e, body) => ...
    }
}
\end{lstlisting}

@@

\subsection{From Interpreter to Staged Interpreter}

\subsubsection{Multi-Stage Programming}
\todo{example, power?}
general idea; LMS framework in scala. MetaOCaml?

>>>>

We give the concrete semantics of \TLang by showing a big-step operational interpreter for it.
But before doing that, let's first see a general interface for interpreters, whether concrete or abstract, 
staged or unstaged. The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Ident},
\texttt{Addr} and etc., which will be instantiated concretely later. The abstract types \texttt{Env} and \texttt{Store}
are required to conform their specifications, \texttt{EnvSpec} and \texttt{StoreSpec}, which define 
the operations such as lookup and update on them, nevertheless the detail of them are omitted.  
The store component is necessary when modeling side effects such as mutations. 
The trait also requires the implementators provide the initial values of environment and store.

@@

Readers may notice an additional abstract type \texttt{R[+T]} is defined in the trait,
which means \texttt{R} represents a higher-kinded type and will used to specify the binding-time, 
i.e., staged or not. Since the LMS framework we are using allows us to specify binding-time purely
based on types, the unstaged higher-kinded type can be simply instantiated as an identity type constructor: 
\texttt{type NoRep[+T] = T}. In section \ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, the address space is infinite thus type \texttt{Int} is used for \texttt{Addr}.
The final returned type by the interpreter \texttt{Ans} is a pair of \texttt{Value} and \texttt{Store}, 
where the values are either closure or numbers.

@@

Then we define evaluation function that dispatches differently for different expression.
As we mentioned before, the type \texttt{R} is concretize as \texttt{NoRep}.
Note that an allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.

\begin{lstlisting}
//TODO: maybe not show all the code
object NoRepConcreteEval extends Concrete {
  type R[+T] = NoRep[T]
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case AOp(op, e1, e2) =>
        val (e1v: NumV, e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v: NumV, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        (evalArith(op, e1v, e2v), e2$\sigma$)
      case Rec(x, f, body) => 
        val $\alpha$ = alloc(x, $\sigma$);  val $\rho$_* = $\rho$ + (x -> $\sigma$)
        val (fv, f$\sigma$) = ev(f, $\rho$_*, $\sigma$)
        val $\sigma$_* = f$\sigma$ + ($\alpha$ -> fv);  ev(body, $\rho$_*, $\sigma$_*)
      case If0(cnd, thn, els) =>
        val (cndv: NumV, cnd$\sigma$) = ev(cnd, $\rho$, $\sigma$)
        if (cndv.i == 0) ev(thn, $\rho$, cnd$\sigma$)
        else ev(els, $\rho$, cnd$\sigma$)
    }
}
\end{lstlisting}

@@

\subsection{From Interpreter to Staged Interpreter} \label{stagedinterp}

\subsubsection{Multi-Stage Programming}
\begin{lstlisting}
val power5 = new DslDriver[Int, Int] {
  def power(b: Rep[Int], x: Int): Rep[Int] =
    if (x == 0) 1 else b * power(b, x-1)
  def snippet(x: Rep[Int]): Rep[Int] = power(x, 5)
}
\end{lstlisting}

Talk about power example. general idea; LMS framework in scala. MetaOCaml?
